---
title: "cardio_HW5"
author: "Joshua Singer"
date: "March 18, 2019"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
library(tidyverse)
library(nlme)
library(car)
library(MASS)
library(lmtest)
library(multcomp)
library(lubridate)
library(gridExtra)
library(nlme)
library(data.table)
library(RCurl)
library(knitr)
library(ggcorrplot)
source('predictgls.R')
source('stdres.gls.R')
source("http://peterhaschke.com/Code/multiplot.R")

```


```{r include=FALSE}
cardio <- read.csv('https://mheaton.byu.edu/Courses/Stat469/Topics/2%20-%20TemporalCorrelation/2%20-%20Longitudinal/HWCaseStudy/Data/Tachycardia.txt', sep=" ", header=TRUE)

```

```{r include=FALSE}
glimpse(cardio)
```

#1: Create exploratory plots of looking at the relationship between HEARTRTE (the response variable) and some of the explanatory variables. Comment on any general relationships you see from the data.
```{r echo=FALSE}

exp_features <- names(cardio[,-c(1, 11)])

#Create scatter plots with every explanatory variable
scat_plots <- map(exp_features, ~ ggplot(cardio, aes_string(x = .x, cardio$HEARTRTE, color = as.factor(cardio$PERIOD))) + geom_point())

#ggplot(cardio, aes_string(x = as.factor(cardio$PERIOD), cardio$HEARTRTE)) + geom_boxplot()

pBox <- ggplot(cardio, aes(as.factor(PERIOD), HEARTRTE)) + geom_boxplot()

grid.arrange(pBox, scat_plots[[3]], scat_plots[[5]], scat_plots[[4]], nrow=2)
```
#2: Fit an independent MLR model with a linear effect of all variables except RANDID and PERIOD. Explore the residuals to see if there is evidence of correlation within a patients from period to period (visit to visit).

```{r echo=FALSE}
cardio.lm <- lm(log(HEARTRTE) ~. -RANDID, data=cardio)

#Check correlation from period to period and the independence assumption
corr_cardio <- matrix(cardio.lm$residuals, ncol = 3, byrow = TRUE) %>% cor()
corr_cardio1 <- data.frame(corr_cardio)
names(corr_cardio1) <- c("Period 1", "Period 2", "Period 3")
ggcorrplot(corr_cardio1, ggtheme = theme_gray, colors = c("#6D9EC1", "white", "#E46726"), lab = TRUE)
```
Above, we can look at the correlation of the residuals to see if there is correlation from period to period. There is a strong correlation across each period with the highest being 0.5 between Period 2 & 3.


#3: To determine an appropriate correlation structure to use, fit a longitudinal MLR model with an AR1, MA1 and general symmetric correlation matrix within each patient but independent across patients. Compare the model fits using AIC (which can be extracted from a gls() object using AIC()).
```{r}
#Model fit with AR1 Correlation Matrix
fit.AR1 <- gls(log(HEARTRTE) ~. -RANDID, data=cardio, correlation=corAR1(form=~1|RANDID), method="ML")

AIC(fit.AR1) #-5915.261

#Model fit with MA1 Correlation Matrix
fit.MA1 <- gls(log(HEARTRTE) ~. -RANDID, data=cardio, correlation=corARMA(form=~1|RANDID, p=0, q=1), method="ML")

AIC(fit.MA1) #-5692.321

#Model fit with General Symmetric Correlation Matrix
fit.Symm <- gls(log(HEARTRTE) ~. -RANDID, data=cardio, correlation=corSymm(form=~1|RANDID), method="ML")

AIC(fit.Symm) #-5993.943
```
The model with the general symmetric correlation matrix has the best fit based on the AIC of -5993.943

#4: Write out your model for analyzing the Tachycardia data in terms of parameters. Explain and interpret any parameters associated with the model.


This is how we will model our analysis using X, $\beta$, and $\epsilon$ with $\beta$ being our matrix of coefficients that correspond to each observation. The dimensions of the $\beta$ matrix is 11 x 1 and the dimensions of the X matrix is n x 11.

$y \sim MV N(X\beta, \sigma^2B)$

$log(y) = X\beta + \epsilon$
In the model the $\beta$ matrix is $(X'X)^{-1}X'y$ where X is the design matrix and y is the vector of response variables. T

$$\mathbf{B} = \left[\begin{array}
{rrr}
R_{1} & \dots & \dots & 0 \\
\vdots & R_{2} & 0 & \vdots\\
\vdots & 0 & \ddots & \vdots \\
0 & \dots & \dots & R_{11}
\end{array}\right]$$



$$\mathbf{R} = \left[\begin{array}
{ccc}
1 & \rho_{12} & \rho_{13}\\
\rho_{21} & 1 & \rho_{23}\\
\rho_{31} & \rho_{32} & 1 \\
\end{array}\right]$$


The B matrix is the matrix we use as a weight for our model's variance. While the R matrix is correlation matrix to test the response based on correlation between periods.

#5: Fit your longitudinal model and validate any assumptions you made to fit the model.
```{r}
fit.Symm <- gls(log(HEARTRTE) ~. -RANDID, data=cardio, correlation=corSymm(form=~1|RANDID), method="ML")

# Linear Assumption 
avPlots(cardio.lm)

# Independent Assuption
sres <- stdres.gls(fit.Symm)
matrix(sres, ncol = 3, byrow = TRUE) %>% cor()

# Normal Assupmetion
ggplot(cardio, aes(x=sres)) + 
  geom_histogram(aes(y=..density..), binwidth=.5, colour="black", fill="white")+
  geom_density(alpha=.2, fill="#FF6666")

ks.test(sres, "pnorm")
# Comments

# Equal Variance
ggplot(data=cardio, mapping=aes(x=(fitted(fit.Symm)), y=sres)) + geom_point()
# Equal Variance assumption looks good. 
```

#6: Is DIABETES a risk factor for Tachycardia? Justify your answer and explain any effect of DIABETES on heart rate (include uncertainty in your conclusions).

Our null hypothesis is $H_0: log(HEARTRTE) = 0$ and our alternative is that $H_a: log(HEARTRTE) > 0$ when $\beta_{diabetes} = 1$.   
```{r}
a <- matrix(c(0,0,0,0,0,0,0,0,1,0,0,0), nrow=1)
hyp.diab <- glht(model = fit.Symm, linfct=a, rhs = 0, alternative="two.sided") %>% summary()
hyp.diab
length(coef(fit.Symm))
length(a)
confint(hyp.diab, .95)
```
Based on this test we fail to reject the null hypothesis the change in heartrate is effected by whether a person has diabetes based on a p-value of 0.503. We are 95% confident that the change in heart rate will only be as high as 3.1%. 


#7: What is the expected difference in heart rate for a female patient with at age 35 who is a smoker vs. an older female of 45 but not a smoker (assume the other characteristics are the same)? What does this say about the effect of smoking?

Our null hypothesis is $H_0: log(HEARTRTE) = 0$ and our alternative is that $H_a: log(HEARTRTE) > 0$ comparing the difference in the change in heart rate between a 35 year old female smoker to a 45 year old female non-smoker.
```{r}
a35 <- matrix(c(0,2,0,35,0,0,1,0,1,0,0,0), nrow=1)
a45 <- matrix(c(0,2,0,45,0,0,0,0,1,0,0,0), nrow=1)

hyp.fitSymm <- glht(model = fit.Symm, linfct=a35-a45, rhs = 0, alternative="two.sided") %>% summary()
hyp.fitSymm

confint(hyp.diab, .95)
confint(hyp.fitSymm, .95)
```
Based on our generalized linear hypothesis test comparing a 35 year old smoker to a 45 year old non-smoker we reject the null hypothesis with a p-value of 4.24e-11. This shows that the difference in log(heartrate) is not 0 and determine that smoking has a dramatic effect (.03756) putting a smoker with a much higher heart rate.

We are 95% confident that the percent change in heart rate lies between 2.6% and 4.8% for women that are smokers versus non-smokers at the age of 35 versus 45 respectively. For our data, we are 95% confident the estimated value is 3.7% change, but are not certain this is true for the entire population.
